\bf Are you normal? A new projection pursuit index to assess a sample against a multivariate null distribution

### 2024-06-25 paper.qmd
ellipse example

@@

### Projection pursuit

### Tours for visualisation

>>>>

```{r include=FALSE}
# Set up chunk for for knitr
knitr::opts_chunk$set(
  fig.width = 5,
  fig.height = 5,
  fig.align = "center",
  out.width = "100%",
  code.line.numbers = FALSE,
  fig.retina = 4,
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  cache = FALSE
)
```

```{r}
#| label: load-libraries
#| warning: false
#| echo: false
library(tourr)
library(mulgar)
library(dplyr)
library(ggplot2)
library(patchwork)
```

```{r}
#| label: plot-theme
theme_set(theme_linedraw() +
   theme(
     aspect.ratio = 1,
     plot.background = element_rect(fill = 'transparent', colour = NA),
     plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
     panel.background = element_rect(fill = 'transparent', 
                                     colour = NA),
     panel.grid.major = element_blank(), 
     panel.grid.minor = element_blank(), 
     axis.title.x = element_blank(), axis.title.y = element_blank(),
     axis.text.x = element_blank(), axis.ticks.x = element_blank(),
     axis.text.y = element_blank(), axis.ticks.y = element_blank(),
     legend.background = element_rect(fill = 'transparent', 
                                      colour = NA),
     legend.key = element_rect(fill = 'transparent', 
                               colour = NA),
     legend.position = "bottom", 
     legend.title = element_blank(), 
     legend.text = element_text(size=4),
     legend.key.height = unit(0.25, 'cm'),
     legend.key.width = unit(0.25, 'cm')
   )
)
interior_annotation <- function(label, position = c(0.92, 0.92)) {
  annotation_custom(grid::textGrob(label = label,
      x = unit(position[1], "npc"), y = unit(position[2], "npc"),
      gp = grid::gpar(cex = 1, col="grey70")))
}
```

@@

To compare a new sample with an existing norm, like a multivariate normal distribution, in higher than two dimensions, we have typically used two samples of points. The norm is represented by points on the surface of a $p$-dimensional ellipse, corresponding to a confidence level. A sample of points uniformly distributed on a $p$-dimensional sphere is generated by

1. Simulating a sample of observations from $N_p(\mu, \Sigma)$. 
2. Transforming each observation to have unit distance from the mean.

```{r}
#| echo: false
#| fig-cap: "Simulating a uniform sample on a sphere involves sampling from a multivariate normal (a) and transforming each observation to have length equal to 1. A confidence ellipse is generated by transforming the sphere relative to a specified variance-covariance matrix (c), and new observations can be visually assessed to be inside or outside by plotting with the ellipse (d)."
set.seed(717)
d1 <- matrix(rnorm(716 * 2), ncol = 2) 
d2 <- t(apply(d1, 1, geozoo:::l2norm_vec))
vc <- matrix(c(1, -0.5, -0.5, 1), ncol=2, byrow=TRUE)
evc <- eigen(vc)
vc2 <- (evc$vectors) %*% diag(sqrt(evc$values)) %*% t(evc$vectors)
d3 <- d2 %*% vc2 |>
  as.data.frame()

d4 <- d3 |>
  bind_rows(data.frame(V1=0.6, V2=0.6)) |>
  bind_rows(data.frame(V1=0.3, V2=0.3)) |>
  mutate(type = c(rep("ci", nrow(d3)), "in", "out"))

d1 <- as.data.frame(d1)
d2 <- as.data.frame(d2)
d3 <- as.data.frame(d3)

p1 <- ggplot(d1, aes(x=V1, y=V2)) +
  geom_point() +
  xlab(expression(x[1])) +
  ylab(expression(x[2])) +
  ggtitle("a. Bivariate standard normal sample") +
  theme(axis.text = element_blank())

p2 <- ggplot(d2, aes(x=V1, y=V2)) +
  geom_point() +
  xlab(expression(x[1])) +
  ylab(expression(x[2])) +
  ggtitle("b. Uniform on a sphere") +
  theme(axis.text = element_blank())

p3 <- ggplot(d3, aes(x=V1, y=V2)) +
  geom_point() +
  xlab(expression(x[1])) +
  ylab(expression(x[2])) +
  ggtitle("c. Points on surface of confidence ellipse.") +
  theme(axis.text = element_blank())

p4 <- ggplot(d4, aes(x=V1, y=V2, shape=type)) +
  geom_point() +
  xlab(expression(x[1])) +
  ylab(expression(x[2])) +
  ggtitle("d. Observations inside and outside confidence ellipse.") +
  theme(legend.position = "none", 
        axis.text = element_blank())

p1 + p2 + p3 + p4 + plot_layout(ncol=2)
```

### 2024-06-25 paper.tex
ellipse example

\subsection{Projection pursuit}\label{projection-pursuit}

\subsection{Tours for visualisation}\label{tours-for-visualisation}

>>>>

To compare a new sample with an existing norm, like a multivariate
normal distribution, in higher than two dimensions, we have typically
used two samples of points. The norm is represented by points on the
surface of a \(p\)-dimensional ellipse, corresponding to a confidence
level. A sample of points uniformly distributed on a \(p\)-dimensional
sphere is generated by

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Simulating a sample of observations from \(N_p(\mu, \Sigma)\).
\item
  Transforming each observation to have unit distance from the mean.
\end{enumerate}

\begin{figure}[H]

{\centering \includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/unnamed-chunk-4-1.pdf}

}

\caption{Simulating a uniform sample on a sphere involves sampling from
a multivariate normal (a) and transforming each observation to have
length equal to 1. A confidence ellipse is generated by transforming the
sphere relative to a specified variance-covariance matrix (c), and new
observations can be visually assessed to be inside or outside by
plotting with the ellipse (d).}

\end{figure}%
